Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2021-07-05 paper/paper.tex
More paper scrawlings

One reason for the success of the study of the retinotectal projection is its
capacity to be experimentally manipulated. In some non-mammalian species, both
the retina and the tectum can be partially ablated, or even physically
reorganised in-vivo, after which axons regrow to restore the order and
function of the system for the individual animal. This manipulability was
exploited in influential work by R. W. Sperry and co-workers during the
mid-twentieth century, leading to Sperry's 1963 summary of
the \emph{chemoaffinity theory}~\cite{sperry_chemoaffinity_1963} which
proposes the existence of morphogenetic gradients that guide axons to their
destination. The chemoaffinity theory was given robust support by the
discovery of the ephrin ligands and their receptors~\cite{cheng_complementary_1995,drescher_vitro_1995}
which have been shown to form into graded expression fields in the
retina~\cite{braisted_graded_1997}, tectum~\cite{braisted_graded_1997,feldheim_genetic_2000} as well as in other
sensory systems, such as the somatosensory
system~\cite{vanderhaeghen_mapping_2000}. The Ephrin ligands have a clear
effect on axonal outgrowth, as shown in in-vitro~\cite{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and
in-vivo~\cite{rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002} studies.
%
% A-P:
%
% EphA and ephrinAs: Anteroposterior axis of retiontectal projection
% (Nasal-Temporal on retina; Anteroposterior on tectum).
%
% RGM (tectum) Neogenin (retina)
%
% D-V:
%
% EphB and ephrin B: Dorsoventral on retina; dorsoventral on tectum.
%
% En-2: Repels temporal growth cones of xenopus ret-tec axons and attracts
% nasal ones. Brunet et all, Nature, 2005
%
% Ryk/Fz (retina) and Wnt3 (tectum) (Flanahan, 2006)
%

@@

@@

@@

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum. In
our model, each retinal ganglion cell projects $n=8$ growth cone agents which
carry a set of 4 receptors, $r$, and 4 ligands, $l$, expressed at levels
determined by the cell's location on the retina. The expression levels of the
receptors and ligands vary with respect to the cell's position on the retinal
surface. Each receptor or ligand varies with respect only to one dimension.
Receptor expression gradients are arranged in orthogonal pairs with the
gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is
opposite to $r_0$, is orthogonal to $r_3$. Ligands are also arranged in
orthogonal pairs, with $l_0$ opposing $r_0$ (as $r_0$ increases, $l_0$
decreases) and $l_1$ orthogonal to $l_0$ and opposing $r_1$.

Because there is evidence that EphA and EphB receptors are expressed in
exponentially increasing
patterns \cite{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\cite{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The tectum expresses the same sets of receptors and ligands, also in
orthogonal pairs of gradients. In this work, we did not consider the effect of
reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression. Although
several studies model tectal ligand expression with exponential functions~\cite{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
X). Correspondingly, we leave open the possibility that ligand expression may
be best modelled by exponential, linear or logarithmic functions.

While we do not
explicitly name $r_0$, $r_1$, etc.~as \emph{EphA}, \emph{EphB}, the suggestion
is that $r$ includes EphA, EphB and Ryk~\cite{schmitt_wntryk_2006} receptors
and that $l$ includes the ephrin-A, ephrin-B and
Wnt3~\cite{schmitt_wntryk_2006} ligands.

@@

This is where I really want to say ``the only interactions in the model are
via forward (ligand-activates-receptor) and reverse
(receptor-activates-ligand) signalling''.

% Hill Equation! https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)

@@

>>>>

One reason for the success of the study of the retinotectal projection is its
capacity to be experimentally manipulated. In some non-mammalian species, both
the retina and the tectum can be partially ablated, or even physically
reorganised in-vivo, after which axons regrow to restore the order and
function of the system for the individual animal. This manipulability was
exploited in influential work by R. W. Sperry and co-workers during the
mid-twentieth century, leading to Sperry's 1963 summary of
the \emph{chemoaffinity theory}~\cite{sperry_chemoaffinity_1963} which
proposes the existence of morphogenetic gradients that guide axons to their
destination. The chemoaffinity theory was given robust support by the
discovery of the ephrin ligands and their receptors~\cite{cheng_complementary_1995,drescher_vitro_1995}
which have been shown to form into graded expression fields in the
retina~\cite{braisted_graded_1997}, tectum~\cite{braisted_graded_1997,feldheim_genetic_2000} as well as in other
sensory systems, such as the somatosensory
system~\cite{vanderhaeghen_mapping_2000}. The Ephrin ligands have a clear
effect on axonal outgrowth, as shown in in-vitro~\cite{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and
in-vivo~\cite{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002} studies.
%
% A-P:
%
% EphA and ephrinAs: Anteroposterior axis of retiontectal projection
% (Nasal-Temporal on retina; Anteroposterior on tectum).
%
% RGM (tectum) Neogenin (retina)
%
% D-V:
%
% EphB and ephrin B: Dorsoventral on retina; dorsoventral on tectum.
%
% En-2: Repels temporal growth cones of xenopus ret-tec axons and attracts
% nasal ones. Brunet et all, Nature, 2005
%
% Ryk/Fz (retina) and Wnt3 (tectum) (Flanahan, 2006)
%

@@

There have been many approaches to modelling the organisation of the
retino-tectal projection. Many of these models are wholly or partially
phenomenological.

@@

Our model was inspired by the agent-based model of \cite{simpson_simple_2011}
and, in common with that work, consists of agents representing axonal growth
cones, originating from a square retina and moving upon a square tectum. Also
in common with the Simpson \& Goodhill model, we model three contributions to
the movement of the growth cones; chemoaffinity, competition and axon-axon
interaction. In our model, chemoaffinity and axon-axon interactions occur via
signal transmission between receptors and ligands which are expressed in
graded patterns on both retina and tectum. This sets our model apart from that
of Simpson \& Goodhill, in which each growth cone moving on the tectum was
always able to determine a vector to its designated target and advance along
that vector. Competition is modelled as a distance based repulsion as
in \cite{simpson_simple_2011}.

@@

Each retinal ganglion cell projects $n=8$ growth cone agents which carry a set
of 4 receptors, $r$, and 4 ligands, $l$, expressed at levels determined by the
cell's location on the retina. The expression levels of the receptors and
ligands vary with respect to the cell's position on the retinal surface. Each
receptor or ligand varies with respect only to one dimension.  Receptor
expression gradients are arranged in orthogonal pairs with the gradient of
$r_0$ being orthogonal to that of $r_1$. $r_2$, whose gradient is opposite to
$r_0$, is orthogonal to $r_3$. Ligands are also arranged in orthogonal pairs,
with $l_0$ opposing $r_0$ (as $r_0$ increases, $l_0$ decreases) and $l_1$
orthogonal to $l_0$ and opposing $r_1$.

Because there is convincing evidence that EphA and EphB receptors are
expressed in exponentially increasing
patterns \cite{reber_relative_2004,feldheim_genetic_2000,brown_topographic_2000,koulakov_stochastic_2004},
we use an exponential form for retinal receptor expressions, in common with
other modelling
studies~\cite{reber_relative_2004,koulakov_stochastic_2004,simpson_simple_2011}.
%
The tectum expresses the same sets of receptors and ligands, also in
orthogonal pairs of gradients. In this work, we did not consider the effect of
reverse signalling from tectal receptors to RGC ligands so we only modelled
tectal ligand expression. Although several studies model tectal ligand
expression with exponential functions~\cite{koulakov_stochastic_2004}, the
experimental evidence for ligand expression is more ambiguous (Supp. Fig
X). Correspondingly, we leave open the possibility that ligand expression may
be modelled by exponential, linear or logarithmic functions.

While we do not explicitly name $r_0$, $r_1$,
etc.~as \emph{EphA}, \emph{EphB}, the suggestion is that $r$ includes EphA,
EphB, Ryk~\cite{schmitt_wntryk_2006} and
Neogenin~\cite{rajagopalan_neogenin_2004} receptors and that $l$ includes the
ephrin-A, ephrin-B, Wnt3~\cite{schmitt_wntryk_2006} and
RGM~\cite{monnier_rgm_2002} ligands, each of which has been shown to play a
role in retino-tectal map formation.

@@

In this model, the signal transmitted when a ligand binds to a receptor on a
growth cone can lead to one of two effects; the signal may induce the growth
cone to climb the gradient of ligand expression, in which case the interaction
is \emph{attractive} or the cone may descend the gradient of ligand expression
if the interaction is \emph{repulsive}.

This is where I really want to say ``the only interactions in the model are
via forward (ligand-activates-receptor) and reverse
(receptor-activates-ligand) signalling''. Somehthing like:

@@

% Discuss the Hill Equation! https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)
